---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# habtools

<!-- badges: start -->
<!-- badges: end -->

The goal of `habtools` is to collate tools for 3D meshes and digital elevation models (DEM) targeted at biologists and ecologists. Tools calculate metrics like surface area, rugosity, fractal dimension, height range, convexity, sphericity, second moments of volume and more. The initial set of tools came from two research papers. Zawada et al. (2019) examined morphology of coral colony laser scans. Torres-Pulliza et al. (2020) examined  complexity of coral reef structure. It is hoped that the number of tools and contributors will grow through time.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("jmadinlab/habtools")
```

## Examples

There are currently two data set accompanying this package. `horseshoe` is a digital elevation model of a reef, and `mcap` is a 3D mesh of a coral growing on a reef. 

The following example calculates height range, rugosity and fractal dimension of a 2 x 2 m plot of reef using the height range method developed in Torres-Pulliza et al. (2020).

```{r example1}
library(habtools)
plot(horseshoe)

# height range
hr(horseshoe, x=-470, y=1266, L=2, plot=TRUE)

# rugosity; note that rugosity will increase with grain (L0)
rg(horseshoe, x=-470, y=1266, L=2, L0=0.25, plot=TRUE)

# fractal dimension is broken into two parts
# 1. Collate height variation in grids of different resolutions
hts <- hvar(horseshoe, x=-470, y=1266, L=2, L0 = 0.25)

# 2. Use the height variation data to calculate fractal dimension
fd(hts)

# Here is the log-log plot with various `methods` superimposed
fd(hts, plot=TRUE)

```

The next example calculates height range, rugosity and fractal dimension for a 3D mesh of a coral colony. Because meshes can have more than one `z` coordinate for a given `xy` (i.e., they have overhangs), we use the cube counting fractal dimension method developed in Zawada et al. (2019).

```{r example2}

# height range
hr(mcap, method="mesh")

# fractal dimension
cubes(mcap, L0 = 0.05, plot=TRUE)

```

You can use the `mesh_to_dem` function and now compare fractal dimension calculated with the two methods. Note that we expect these to be somewhat different is only because DEMs do not include overhangs (i.e., they have one `z` per `xy` pair; or as some people say, they are 2.5D, not 3D).

```{r example3}
dem <- mesh_to_dem(mcap, res = 0.01)
plot(dem)
L <- min(dim(dem)[c(1,2)]) * res(dem)[1]

hts <- hvar(dem, x = xmin(dem), y = ymin(dem), L = L, L0 = 0.05)
fd(hts, plot = TRUE)

```

