---
title: "Complexity metrics for DEMs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complexity metrics for DEMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(habtools)
library(raster)
library(dplyr)
library(ggplot2)
```

In this vignette, we go over the basic steps to calculate standard complexity metrics for digital elevation models (DEMs).
In this example, we use a DEM of 8x8m.

```{r}
# Load a raster
dem <- horseshoe
res(dem)
plot(dem)
# The resolution is 0.002m
```
Let's say we want complexity metrics for a certain square around xy-coordinates of size 2m.
```{r}
dem1 <- crop_dem(horseshoe, -466, 1269, L = 2, plot = TRUE)
plot(dem1)
```

## Height range

```{r}
hr(dem1)
```

## Rugosity

```{r}
# Height variation method
rg(dem1, L0 = 0.1, method = "hvar", parallel = FALSE) # Parallel = TRUE enables parallel processing using multiple cores to speed up the calculations using the height variation method. Only use this if you have a powerful computer with at least four cores. 
           
# Area method
rg(dem1, L0 = 0.1, method = "area")
```

## Fractal dimension

```{r}
# Height variation method
fd(dem1, lvec = c(0.125, 0.25, 0.5, 1, 2), plot = T, method = "hvar")
# Area method
fd(dem1, lvec = c(0.05, 0.1, 0.25, 0.5), plot = T, method = "area")
```

## Multiple metrics

```{r}
rdh(dem1, lvec = c(0.0625, 0.125, 0.25, 0.5, 1, 2), method = "hvar")
rdh(dem1, lvec =  c(0.05, 0.1, 0.25, 0.5), method = "area")
```

## Multiple squares
The function `split_dem()` allows you to subdivide a DEM into equal parts. Let's divide the square into 2x2m squares. 
```{r}
dem_list <- split_dem(dem, size = 2)
# calculate one metric for all squares
sapply(dem_list, hr)

# calculate multiple metrics
data_rdh <- lapply(dem_list, rdh,  method = "hvar", lvec = c(0.0625, 0.125, 0.25, 0.5, 1, 2)) %>%
  bind_rows()

```

```{r}
ggplot(data_rdh) +
  geom_point(aes(x = R, y = H, color = D, size = D)) +
  theme_classic()
```


## Rules of thumb

- all res have to be same across dems
- smallest lvec hvar method at least 10x res
- extent (heighest of lvec) 1 or 2 orders of magnitudes compared to smallest lvec (TEST THIS!) (or scales of interest)
- lvec needs to be the same across all calculations to make comparisons


